A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-06-15 tex/BiSig.tex
[ tex ] Emphaise generality and practicality of our theory

  Motivated by this challenge, this paper contributes to the development of a general and practical theory of type checking and inference.
  We present a \emph{theory} of bidirectional typing, which had only been a \emph{design discipline}, for synthesising types to handle \emph{syntax-directed} bidirectional type systems of simple types that can be specified by a bidirectional binding signature.
  Here, \emph{type synthesis} refers to the mutual use of type checking and inference to synthesise a type from programs possibly with type annotations.

  To establish the correctness and effectiveness of our proposal, we formalize our theory in vanilla \Agda.
  We prove logical decidability of type synthesis, subsuming soundness and completeness of type synthesis algorithm.
  By Curry-Howard correspondence, our theorems are actually correct-by-construction synthesisers.
  We also discuss another essential characteristic of bidirectional typing called \emph{annotability} and propose a related but simpler and practical notion called \emph{bidirectionalisation}.
  Combining with decidability of type synthesis, we derive the trichotomy of raw terms by type synthesis.
  Put differently, the trichotomy is a type synthesiser which distinguishes programs with missing type annotations from programs that cannot be checked.

  In summary, this paper represents a step towards a general and practical theory of type checking and inference.
  By formalizing our theory of bidirectional type synthesise in \Agda, we provide a framework capable of handling bidirectional type systems of simple types.
  Our work addresses the problem with existing methodologies, setting the stage for further advances towards a general theory of type synthesis.
\end{abstract}

>>>>

  Motivated by this challenge, this paper contributes to the development of a general and practical theory of type checking and inference.
  First, we present a \emph{general} theory of bidirectional typing, which had only been a design discipline, to synthesise types for \emph{syntax-directed} bidirectional type systems of \emph{simple types}.
  Here, \emph{(bidirectional) type synthesis} refers to the mutual use of type checking and inference to synthesise the type of a programs possibly with type annotations.
  Second, our theory is \emph{practical} not only because we formalize our theory in \Agda but also we have formulated our results from the practical aspect to provide useful computational meaning.
  Specifically, we prove \emph{logical decidability} of type synthesis, subsuming soundness, completeness, and decidability of type synthesis algorithm; by Curry-Howard correspondence, our proof is a correct-by-construction synthesiser.
  Also we discuss an essential characteristic of bidirectional typing called \emph{annotability} and propose a related but simpler and practical replacement called \emph{bidirectionalisation}.
  Combining with decidability of type synthesis, we derive the trichotomy of raw terms whose proof is a type synthesiser which distinguishes raw terms with missing type annotations from those that cannot be checked.

  In short, this paper is a first step towards a general and practical theory of type checking and inference.
  By formalizing our theory of bidirectional type synthesise in \Agda, we provide a framework capable of handling bidirectional type systems of simple types.
  Our work addresses the problem with existing methodologies, setting the stage for further advances towards a general theory of type synthesis.
\end{abstract}
